---
title: "Sensors"
author: "Meghasyam Tummalacherla"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sensors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`mhealthtools` offers a set of tools to extract features from some of the sensors present in almost all of the modern day phones. Here we present examples from accelerometer and gyroscope data.

Let's load the package first
```{r mhealthtools}
library(mhealthtools)
```

```{r other libraries, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, prompt=FALSE}
library(dplyr)
```

## Accelerometer

#### Data format and Sample data

`mhealthtools` already comes preloaded with sample accelerometer data. You can access the data through `data(accelerometer_data)` or just `accelerometer_data`. The first five rows are shown below, to give you an idea of the structure of the data (this is the input format to the functions that extract accelerometer features). Where `t` is the timestamp; `x`, `y` and `z` are the X, Y and Z- axis respectively.

```{r accel data, echo = FALSE}
head(accelerometer_data, 5)
```
*To use the default feature extraction pipelines you need to get the data in the proper format -- as that of the sample data. You can also define your own features or feed in a custom model to extract features in `mhealthtools` as described in [Extending mhealthtools](extending.html) under the "Sensor modules" section header.*

Let's plot to see how the accelerometer data actually looks like 

```{r accel plots, echo=FALSE, fig.width=8, fig.height=4}
library(ggplot2)
a <- accelerometer_data
a$t <- a$t - a$t[1]
a <- tidyr::gather(a, 'axis' , 'value', -t)
ggplot(a, aes(x = t, y= value)) + geom_line() + 
  facet_wrap(~axis, dir = 'v', strip.position = 'right') + ylim(c(-0.025, 0.025))
```
We are looking at 10s of data sampled at a high enough sampling rate (for accelerometer data) of 100Hz. The data looks noisy, and looks like it could make good use of some pre-processing like frequency filtering, maybe subsetting the time (time filtering) -- all of which `mhealthtools` conveniently offers.

#### Extracting features (using pre-existing pipeline)

Let's take a look at the default feature extraction pipeline, whose features are all classical signal processing features based on time domain, frequency domain and the energy bands of the frequency spectrum.

```{r accel default features, warning=FALSE, error=FALSE, prompt=FALSE}
accel_features <- accelerometer_features(accelerometer_data)
# using default parameters
```
The output file is a list containing features extracted using default functions (`$extracted_features`), features calculated using models (`$model_features`) and an element for error (`$error`). Let's look at the `$extracted_features` to see how the default features look like

```{r show accel default features}
dplyr::glimpse(accel_features$extracted_features)
```

`accelerometer_features` also offers multiple parameter options. The following examples are an effort to make the user aware of the options. 

In the following example, we change the window length (to 128, with NULL being the default - i.e no windowing), having the 50% overlap between consecutive windows for the same `accelerometer_data` used before

```{r, warning=FALSE, error=FALSE, prompt=FALSE}
accel_features <- accelerometer_features(
  accelerometer_data,
  window_length = 128,
  window_overlap = 0.5)
# Consider a window length of 128 samples,
# 50% overlap between consecutive windows
```

```{r}
dplyr::glimpse(accel_features$extracted_features)
```
Notice how we now have three extra columns in the feature extraction output namely, `window`, `window_start_time` and `window_stop_time` - which indicate the number of the window in the signal (the first window is window 1, the next consecutive window is window 2 and so on), and the start and stop time respectively of that window.

There are many other possible use cases to calculate the accelerometer features. The `time_filter` parameter decides the time window to be operated on in the accelerometer signal. The `window_overlap` parameter decides the amount of overlap between consecutive windows. The `frequency_filter` parameter specifies the frequency range to keep for the analysis.

```{r, warning=FALSE, error=FALSE, prompt=FALSE, eval=FALSE}
accel_features <- accelerometer_features(accelerometer_data, 
                                         time_filter = c(2,5))
# Consider the time ranges between 2 and 5s only
```

```{r, warning=FALSE, error=FALSE, prompt=FALSE, eval=FALSE}
accel_features <- accelerometer_features(accelerometer_data, 
                                         frequency_filter = c(4,16))
# Consider the frequencies only from 4Hz to 16Hz
```

There are also advanced processing techniques that you can apply to the signal like detrending (using loess), or decompose the signal into Intrinsic Mode Functions (IMF) using the Hilbert Huang Transform. `detrend` and `IMF` respectively are the parameters that refer to these techniques.

```{r, warning=FALSE, error=FALSE, prompt=FALSE, eval=FALSE}
accel_features <- accelerometer_features(accelerometer_data, 
                                         IMF = 2)
# Consider decomposing the signal into 2 IMFs and then running both the IMFs through
# the default feature extraction pipeline
```

Please read the function documentation `?accelerometer_features` for more info.

## Gyroscope

#### Data format and Sample data

`mhealthtools` already comes preloaded with sample gyroscope data. You can access the data through `data(gyroscope_data)` or `mhealthtools::gyroscope_data`. The format is similar to the accelerometer data described above.

```{r gyro data, echo = FALSE}
head(gyroscope_data,5)
```

Let's plot to see how the gyroscope data actually looks like 

```{r gyro plots, echo=FALSE, fig.width=8, fig.height=4}
library(ggplot2)
a <- gyroscope_data
a$t <- a$t - a$t[1]
a <- tidyr::gather(a, 'axis' , 'value', -t)
ggplot(a, aes(x = t, y= value)) + geom_line() +
  facet_wrap(~axis, dir = 'v', strip.position = 'right') + ylim(c(-0.025, 0.025))
```

#### Extracting features (using pre-existing pipeline)

The format of the gyroscope feature extraction pipeline is similar to that of the accelerometer feature extraction pipeline mentioned above. Let's take a look at the default features.

```{r gyro default features, warning=FALSE, error=FALSE, prompt=FALSE}
gyro_features <- gyroscope_features(gyroscope_data)
# using default parameters
```

```{r show gyro default features}
dplyr::glimpse(gyro_features$extracted_features)
```

`gyroscope_features` also offers multiple parameter options. The following examples are an effort to make the user aware of the options. Please look at the `Accelerometer` section above for a detailed description.

```{r, warning=FALSE, error=FALSE, prompt=FALSE}
gyro_features <- gyroscope_features(
  gyroscope_data,
  window_length = 128,
  window_overlap = 0.5)
# Consider a window length of 128 samples,
# 50% overlap between consecutive windows
```

```{r}
dplyr::glimpse(gyro_features$extracted_features)
```

```{r, warning=FALSE, error=FALSE, prompt=FALSE, eval=FALSE}
gyro_features <- gyroscope_features(gyroscope_data, 
                                         time_filter = c(2,5))
# Consider the time ranges between 2 and 5s only
```

```{r, warning=FALSE, error=FALSE, prompt=FALSE, eval=FALSE}
gyro_features <- gyroscope_features(gyroscope_data, 
                                         frequency_filter = c(4,16))
# Consider the frequencies only from 4Hz to 16Hz
```

```{r, warning=FALSE, error=FALSE, prompt=FALSE, eval=FALSE}
gyro_features <- gyroscope_features(gyroscope_data, 
                                         IMF = 2)
# Consider decomposing the signal into 2 IMFs and then running both the IMFs through
# the default feature extraction pipeline
```

Please read the function documentation `?gyroscope_features` for more info.

## Screen (Tap data)

For more information on how this data was collected please look into the Parkinson mpower study conducted by Sage Bionetworks (https://parkinsonmpower.org/)

#### Data format and Sample data

`mhealthtools` already comes preloaded with sample tapping data. You can access the data through `data(tap_data)` or `mhealthtools::tap_data`. The first five rows are shown below, to give you an idea of the structure of the data (this is the input format to the functions that extract gyroscope features). Where `t` is the timestamp; `x` and `y` are the co-ordinates of the tap, and `buttonid` is to indicate whether the user tapped the left or the right button.

```{r, echo = FALSE}
head(tap_data,5)
```

Let's look at a plot of the `x` and `y` co-ordinates of the tap data against `t`

```{r, echo=FALSE, fig.width=8, fig.height=4}
library(ggplot2)
a <- tap_data
a <- tidyr::gather(a %>% dplyr::select(-buttonid), 'axis' , 'value', -t)
ggplot(a, aes(x = t, y= value)) + geom_line() + facet_wrap(~axis, dir = 'v', strip.position = 'right')
```

#### Extracting features (using pre-existing pipeline)
To use the existing feature set (you can define your own features or feed in your own model to extract features in `mhealthtools`!!). Note: You need to get the data in the proper format (as that of the sample data) to use the feature extraction pipeline.

```{r, warning=FALSE, error=FALSE, message=FALSE, echo = FALSE}
library(mhealthtools)
library(dplyr)
```
```{r, warning=FALSE, error=FALSE, prompt=FALSE}
tap_features <- get_tapping_features(tap_data)
```
```{r}
dplyr::glimpse(tap_features)
```

You can also specify the threshold for the intertap distance in the x-axis using the `depress_threshold` parameter of the get_tapping_features function

```{r, warning=FALSE, error=FALSE, prompt=FALSE}
tap_features <- get_tapping_features(tap_data, depress_threshold = 10)
# Changing the threshold for intertap distance in x-axis, depress_threshold to 10 (default is 20)
```
```{r}
dplyr::glimpse(tap_features)
```

Please read the function documentation `?get_tapping_features` for more info.